Appendix S3. The relative importance of evolution in populations and communities

### 2024-11-30 manuscript/Appendix_S3_v01.Rmd
Updated mscript, all appendices, all figures.

---
title: "Appendix S3. Hypotheses and tests for increased importance of evolutionary drivers for species and community dynamics"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(Hmsc)
library(ggplot2)
library(tidyr)
library(coda)
library(patchwork)
library(jtools)
library(bayesplot)
library(reshape2)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

# Background

First I establish a metacommunity simulation scenario where environment, spatial random effects, and year-based Random Effects (species interactions) all play a role in driving species abundances. I then add in trait evolution, and look at its relative importance.

# Model for population growth, competition, environmental change, and evolution

The growth equation for each species is:

@@

We simulate population growth under particular parameter values, to generate population dynamics where species interactions, environment, and trait evolution all drive population dynamics.

```{r}
set.seed(42)
# Case 1: Weaker interactions, stronger environment
# Simulate initial species population growth with environment fluctuations
N1.0 <- 10
N2.0 <- 10
alpha.11 <- 0.01
alpha.22 <- 0.01
alpha.12 <- 0.005
alpha.21 <- 0.01
E.0 <- 0.8
x1.0 <- 0.1
x2.0 <- 0.8
P <- 1
w <- 2
Wmax <- 2
h2 <- 1
k <- (w + (1 - h2) * P)/(P + w)

# model function
disc_LV_evol <- function(N1.0,N2.0,alpha.11,alpha.22,alpha.12,alpha.21,E,x1.0,x2.0,P,w,Wmax,h2) {
  What <- Wmax*sqrt(w/(P+w))
  r1 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x1.0))^2)/(2*(P+w)))
  Nt1 <- (r1*N1.0) / (1+alpha.11*N1.0+alpha.12*N2.0)
  r2 <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E-x2.0))^2)/(2*(P+w)))
  Nt2 <- (r2*N2.0) / (1+alpha.22*N2.0+alpha.21*N1.0)
  return(c(Nt1,Nt2,r1,r2))
}
# Simulation of model for t time steps
t <- 40
N <- array(NA,dim=c(t,2))
N <- as.data.frame(N)
colnames(N) <- c("N1","N2")
x <- array(NA,dim=c(t,2))
x <- as.data.frame(x)
colnames(x) <- c("x1","x2")
r <- array(NA,dim=c(t,2))
r <- as.data.frame(r)
colnames(r) <- c("r1","r2")
N$N1[1] <- N1.0
N$N2[1] <- N2.0
x$x1[1] <- x1.0
x$x2[1] <- x2.0
What <- Wmax*sqrt(w/(P+w))
r$r1[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
r$r2[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
E <- rep(NA, t)
E[1] <- E.0
for (i in 2:t) {
  res <- disc_LV_evol(N1.0=N[i-1,1],N2.0=N[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=x[i-1,1],x2=x[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
  N$N1[i] <- res[1]
  N$N2[i] <- res[2]
  r$r1[i] <- res[3]
  r$r2[i] <- res[4]
  # trait change
  d1 <- E[i-1] - x[i-1,1]
  d2 <- E[i-1] - x[i-1,2]
  d1.1 <- k*d1
  d2.1 <- k*d2
  x$x1[i] <- E[i-1] - d1.1
  x$x2[i] <- E[i-1] - d2.1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1,0,0.05))
}
```
```{r}
# Plot simulation: ggplot
N$time <- 1:t
dat <- melt(N, id.vars="time")
ggplot2::ggplot(dat, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = (1/alpha.11),linetype = "dashed", color = "gray")
```

We can check the impact evolution had for this system by re-running dynamics with $h^2 = 0$:

```{r echo = FALSE}
h2 <- 0
k <- (w+(1-h2)*P)/(P+w)
# Simulation of model for t time steps
Na <- array(NA,dim=c(t,2))
Na <- as.data.frame(Na)
colnames(Na) <- c("N1","N2")
xa <- array(NA,dim=c(t,2))
xa <- as.data.frame(xa)
colnames(xa) <- c("x1","x2")
ra <- array(NA,dim=c(t,2))
ra <- as.data.frame(ra)
colnames(ra) <- c("r1","r2")
Na$N1[1] <- N1.0
Na$N2[1] <- N2.0
xa$x1[1] <- x1.0
xa$x2[1] <- x2.0
What <- Wmax*sqrt(w/(P+w))
ra$r1[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x1.0))^2)/(2*(P+w))))
ra$r2[1] <- What*(exp((-(((w+(1-h2)*P)/(P+w))*(E.0-x2.0))^2)/(2*(P+w))))
for (i in 2:t) {
  res <- disc_LV_evol(N1.0=Na[i-1,1],N2.0=Na[i-1,2],alpha.11=alpha.11,alpha.22=alpha.22,alpha.12=alpha.12,alpha.21=alpha.21,E=E[i-1],x1=xa[i-1,1],x2=xa[i-1,2],P=P,w=w,Wmax=Wmax,h2=h2)
  Na$N1[i] <- res[1]
  Na$N2[i] <- res[2]
  ra$r1[i] <- res[3]
  ra$r2[i] <- res[4]
  # trait change
  d1 <- E[i-1] - xa[i-1,1]
  d2 <- E[i-1] - xa[i-1,2]
  d1.1 <- k*d1
  d2.1 <- k*d2
  xa$x1[i] <- E[i-1] - d1.1
  xa$x2[i] <- E[i-1] - d2.1
}
# Plot simulation: ggplot
Na$time <- 1:t
data <- melt(Na, id.vars="time")
ggplot2::ggplot(data, aes(time, value, col=variable)) + geom_point() + geom_hline(yintercept = (1/alpha.11),linetype = "dashed", color = "gray")
```

# Hypotheses for evolution as a driver of species abundances

The growth equation is not linear, yet we fit abundance data to a hierarchical model with linear predictors because in most natural systems, a single theoretical growth model can't adequately capture the complexity in a given system. Linear models don't capture mechanistic relationships, but instead correlative. They are useful for inferring direction and magnitude of effect sizes. In Appendix S1, we explore how non-linear dynamics are captured when fit to a linear model. Here, we explore different forms to encode evolution in the trait $x$ that determines the organism's fitness in this system.

## H1. Raw trait value drives species abundances

```{r}
# Plot simulation: ggplot
dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
colnames(dat) <- c("N1","N2","x1","x2")
dat$time <- 1:t
df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
colnames(df) <- c("Nt1","Nt2","xt1","xt2")
# Plot
p1 <- ggplot2::ggplot(df, aes(xt1, Nt1)) + geom_point()
p2 <- ggplot2::ggplot(df, aes(xt2, Nt1)) + geom_point()
p3 <- ggplot2::ggplot(df, aes(xt1, Nt2)) + geom_point()
p4 <- ggplot2::ggplot(df, aes(xt2, Nt2)) + geom_point()
p1 + p2 + p3 + p4
```

We do not hypothesize that the raw trait value is the best predictor of species abundances, because the impacts of this evolved trait depend on the context of the environment.

## H2. Absolute value of trait distance from optimum drives species abundances

```{r}
# Plot simulation: ggplot
df$E <- E[2:t]
df$abs_Ex1 <- abs(df$E - df$xt1)
df$abs_Ex2 <- abs(df$E - df$xt2)
# Plot
p1 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt1)) + geom_point()
p2 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt1)) + geom_point()
p3 <- ggplot2::ggplot(df, aes(abs_Ex1, Nt2)) + geom_point()
p4 <- ggplot2::ggplot(df, aes(abs_Ex2, Nt2)) + geom_point()
p1 + p2 + p3 + p4
```

We hypothesize that this represents the true impact on species abundances, via the direct impact this measure has on fitness and consequent population growth. However, we also acknowledge this is likely difficult to measure in most empirical systems.

## H3. Absolute value of trait change from one time to the next drives species abundances

We now onsider the magnitude of trait change from one time step to the next as a driver for species abundances.

```{r}
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
# Plot
p1 <- ggplot2::ggplot(df, aes(dx1, Nt1)) + geom_point()
p2 <- ggplot2::ggplot(df, aes(dx2, Nt1)) + geom_point()
p3 <- ggplot2::ggplot(df, aes(dx1, Nt2)) + geom_point()
p4 <- ggplot2::ggplot(df, aes(dx2, Nt2)) + geom_point()
p1 + p2 + p3 + p4
```

The measures correlate highly with one another, as the amount of trait change (the amount of evolution) is determined by the distance to the optimum trait value (given that P and w remain constant in the simulation).

```{r}
cor(df$dx1, df$abs_Ex1)
cor(df$dx2, df$abs_Ex2)
# Plot
p1 <- ggplot2::ggplot(df, aes(abs_Ex1,dx1)) + geom_point()
p2 <- ggplot2::ggplot(df, aes(abs_Ex2,dx2)) + geom_point()
p1 + p2
```

# Statistical model for evolution as a driver of species abundances

We hypothesize that \( \lvert E_t - x_{i,t} \rvert \) will be the best predictor of species abundances, and that \( \lvert x_{t+1} - x_{i,t} \rvert \) will have similar explanatory power.

```{r}
# prepare data in HMSC format
dat <- as.data.frame(cbind(log(N$N1),log(N$N2),x$x1,x$x2))
colnames(dat) <- c("N1","N2","x1","x2")
dat$time <- 1:t
df <- data.frame(cbind(dat$N1[2:t],dat$N2[2:t],dat$x1[2:t],dat$x2[2:t]))
colnames(df) <- c("Nt1","Nt2","xt1","xt2")
df$dx1 <- abs(dat$x1[2:t] - dat$x1[1:(t-1)])
df$dx2 <- abs(dat$x2[2:t] - dat$x2[1:(t-1)])
# Y matrix
Y <- as.matrix(cbind(df$Nt1,df$Nt2))
# X matrix
XData <- data.frame(cbind(dat$N1[1:(t-1)],dat$N2[1:(t-1)]),E[1:(t-1)],E[1:(t-1)]^2,dat$x1[1:(t-1)],dat$x2[1:(t-1)],abs(dat$x1[2:t] - dat$x1[1:(t-1)]),abs(dat$x2[2:t] - dat$x2[1:(t-1)]),abs(E[1:(t-1)] - dat$x1[1:(t-1)]),abs(E[1:(t-1)] - dat$x2[1:(t-1)]))
colnames(XData) <- c("n1","n2","E","Esq","x1","x2","dx1","dx2","dEx1","dEx2")
# Prepare HMSC model
studyDesign = data.frame(sample = as.factor(1:(t-1)))
rL = HmscRandomLevel(units = studyDesign$sample)
# Design and fit 4 alternative HMSC models
models = list()
for(i in 1:4){
  XFormula = switch(i, ~1, ~E+Esq+x1+x2, ~E+Esq+dEx1+dEx2, ~E+Esq+dx1+dx2)
  m = Hmsc(Y = Y, XData = XData, XFormula = XFormula, studyDesign = studyDesign, ranLevels = list(sample = rL))
  models[[i]] = m
}
# Bayesian model parameters
nChains <- 2
thin <- 5
samples <- 2000
transient <- 100*thin
verbose <- 500*thin
for(i in 1:4){
  models[[i]] = sampleMcmc(models[[i]],thin=thin,sample=samples,transient=transient,nChains=nChains,verbose=verbose)
}
```

We check the explanatory and predictive power of each model using cross-validation.

```{r}
# After Ovaskainen & Abrego 2020, Chapter 7
partition = createPartition(m,nfolds=2,column="sample")
partition.sp = c(1,2)
result = matrix(NA,nrow=3,ncol=4)
for(i in 1:4){
  m = models[[i]]
  # Explanatory power
  preds = computePredictedValues(m)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[1,i] = mean(MF$R2)
  # Predictive power based on cross-validation
  preds = computePredictedValues(m,partition=partition)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[2,i] <- mean(MF$R2)
  # Predictive power based on conditional cross-validation
  preds = computePredictedValues(m,partition=partition,partition.sp=partition.sp,mcmcStep=100)
  MF = evaluateModelFit(hM=m,predY=preds)
  result[3,i] = mean(MF$R2)
}
```

```{r}
colnames(result) <- c("null","xt","dEx","dx")
rownames(result) <- c("R2_expl","R2_pred","R2_pred_cross")
result
```

An important conclusion from this analysis is that the \( \lvert E_t - x_{i,t} \rvert \) and \( \lvert x_{t+1} - x_{i,t} \rvert \) do have similar explanatory power, and that - even though the species traits were the stronger predictor of species abundances - the model with total change in trait value can be used to assess the question "How does evolution impact species abundances?"

```{r}
m.post.hmsc <- convertToCodaObject(models[[4]])
bayesplot::mcmc_trace(m.post.hmsc$Beta)
bayesplot::mcmc_areas(m.post.hmsc$Beta,area_method = c("equal height"))
```

```{r}
# Bayesian estimates
cbind(summary(m.post.hmsc$Beta)$statistics[1:10,1],summary(m.post.hmsc$Beta)$quantiles[,c(1,5)])
```

```{r}
OmegaCor = computeAssociations(models[[4]])
supportLevel = 0.7
toPlot = ((OmegaCor[[1]]$support>supportLevel)
+ (OmegaCor[[1]]$support<(1-supportLevel))>0)*OmegaCor[[1]]$mean
corrplot::corrplot(toPlot, method = "color",col = colorRampPalette(c("blue","white","red"))(200),title = paste("random effect level:", models[[4]]$rLNames[1]), mar=c(0,0,1,0))
OmegaCor
```
We can also look at a gradient plot for the effect of evolution in Species 1 for abundance in Species 1 and Species 2.

```{r}
Gradient <- constructGradient(models[[4]],focalVariable="dx1",non.focalVariables=list(E=list(2),Esq=list(2),dx2=list(2)),ngrid=39)
Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
predY <- predict(models[[4]],XData=Gradient$XDataNew,expected=TRUE)
a <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx1_t",ylab="predicted N1_t+1")
b <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx1_t",ylab="predicted N2_t+1")

postBeta = getPostEstimate(models[[4]], parName = "Beta")
plotBeta(models[[4]], post = postBeta, param = "Mean", supportLevel = 0.7)
```
We repeat this plot for the effect of evolution in Species 2 for abundance in Species 1 and Species 2.

```{r}
Gradient <- constructGradient(models[[4]],focalVariable="dx2",non.focalVariables=list(E=list(2),Esq=list(2),dx1=list(2)),ngrid=39)
Gradient$XDataNew$Esq <- Gradient$XDataNew$E^2
predY <- predict(models[[4]],XData=Gradient$XDataNew,expected=TRUE)
a <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=1,main="",xlab="dx2_t",ylab="predicted N1_t+1")
b <- plotGradient(models[[4]],Gradient,pred=predY,showData=T,measure="Y",index=2,main="",xlab="dx2_t",ylab="predicted N2_t+1")
```
To evaluate the relative importance of evolution for species abundances, we use variation partition:

>>>>

---
title: "Appendix S3. The relative importance of evolution in populations and communities"
author:
  - Jelena H. Pantel^[Laboratoire Chrono-environnement,UMR 6249 CNRS-UFC, 16 Route de Gray, 25030 Besançon cedex, France, jelena.pantel@univ-fcomte.fr]
  - Ruben J. Hermann^[University of Duisburg-Essen, Universitätsstraße 5, 45141 Essen, Germany, ruben.hermann@uni-due.de]
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    extra_dependencies: ["float"]
  bookdown::html_document2: default
  word_document:
    reference_docx: docx_template.docx
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{Author Pantel and Hermann}
- \fancyhead[R]{Detection of ecoevo dynamics}
- \usepackage{lineno}
- \linenumbers
link-citations: yes
linkcolor: blue
csl: the-american-naturalist.csl
bibliography: refs.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
```{r setup,echo=FALSE,message=FALSE}
library(ecoevor)
library(ggplot2)
library(reshape2)
library(Hmsc)
library(bayesplot)
```
```{r,echo=FALSE,message=FALSE,warning=FALSE}
# Wrap long lines
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE,out.width="50%")
```

# Logistic growth and trait evolution

The goal of this analysis is to show that a hierarchical Bayesian linear model (HMSC) can successfully estimate the relative importance of different drivers of population dynamics, and to evaluate how the importance of evolution for population dynamics changes with heritability $h^2$ level. This analysis supports the findings for the **Evolution in metacommunities and HMSC** section in the main text. We create simulations of evolutionary rescue in a population, where adaptive evolution rescues a population from extinction, at a speed that depends on the heritability level in the system. Population dynamics are modelled as:

@@

where \( \hat{W} \) is calculated as \( \hat{W}=W_{max}\sqrt(\frac{w}{P+w}) \), \(W_{max}\) is the species' maximum per-capita growth rate, *w* is the width of the Gaussian fitness function (which determines the strength of selection, as increasing values indicate a weaker reduction in fitness with distance from optimum trait value), *P* is the width of the distribution of the phenotype *x*, and $h^2$ is the heritability of the trait *x*. For the simulation we use \(W_{max}\) = 2, *P* = 1, and *w* = 2.

The change in the average trait value each time step is given by:

$$ d_{i,t+1} = k d_{i,t} $$
where \( k = \frac{w + (1-h^2) P}{w + P} \) and \( d_{i,t} = E_t - x_{i,t} \).

## Population dynamics simulation

We simulate population dynamics for 1 species, keeping all conditions constant except for heritability, which is evaluated at $h^2 = 0, 0.01, 0.02, 0.03, 0.04, 0.05, 0.075, 0.1, 0.4, 0.9, 1$. The simulation code for only $h^2 = 0$ is shown.

```{r}
### 1 species, h2=0 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- N
gdat$time <- 1:t
gdat$h2 <- h2
gdat$x1 <- x$x1
gdat$E <- E
```
```{r echo=FALSE}
### 1 species, h2=0.01 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.01
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r echo=FALSE}
### 1 species, h2=0.02 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.02
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r echo=FALSE}
### 1 species, h2=0.03 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.03
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r echo=FALSE}
### 1 species, h2=0.04 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.04
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r echo=FALSE}
### 1 species, h2=0.05 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.05
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r echo=FALSE}
### 1 species, h2=0.075 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.075
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
gdat <- rbind(gdat,data.frame(N,time=1:t,h2=rep(h2,t),x1=x$x1,E=E))
```
```{r echo=FALSE}
### 1 species, h2=0.1 ###
# Initial conditions
N0 <- 10
alpha.11 <- 0.01
alpha <- matrix(alpha.11,byrow=FALSE)
E.0 <- 0.8
P <- 1
w <- 0.5
# Draw initial trait value using degree of initial maladaptation
B0 <- 1
d0 = sqrt(B0*(w+P))
x.0 = E.0-d0
Wmax <- 2
h2 <- 0.1
k <- (w + (1 - h2) * P)/(P + w)
# Simulation of model for t time steps
# Simulation of model for t time steps
t <- 60
N <- array(NA,dim=c(t,length(N0)))
N <- as.data.frame(N)
colnames(N) <- paste0("N",1:length(N0))
N[1,] <- N0
E <- rep(NA, t)
E[1] <- E.0
x <- array(NA,dim=c(t,length(N0)))
x <- as.data.frame(x)
colnames(x) <- paste0("x",1:length(N0))
x[1,] <- x.0
r <- array(NA,dim=c(t,length(N0)))
r <- as.data.frame(r)
colnames(r) <- paste0("r",1:length(N0))
What <- Wmax*sqrt(w/(P+w))
r[1,] <- What*exp((-(((w+(1-h2)*P)/(P+w))*(E[1]-x[1,]))^2)/(2*(P+w)))
for (i in 2:t) {
  res <- disc_LV_evol(N0=N[i-1,],alpha=alpha,E=E[i-1],x=x[i-1,],P=P,w=w,Wmax=Wmax,h2=h2)
  N[i,] <- res$Nt1
  r[i,] <- res$r
  # trait change
  d <- E[i-1] - x[i-1,]
  d1 <- k*d
  x[i,] <- E[i-1] - d1
  # environmental change
  E[i] <- E[i-1] + abs(rnorm(1, 0, 0))
}
# Arrange to save across scenarios
